---
title: "Mathematical Foundations of the Quadratic-Form Permutation Saddlepoint"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mathematical Foundations of the Quadratic-Form Permutation Saddlepoint}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

This memo explains how `fastperm` computes the tail probability of a
quadratic-form test statistic under a within-stratum permutation null, why each
step is built the way it is, and where each decision lives in the code and in the
test suite. It is written to be read start to finish: the plain idea first, a
small concrete case next, then the notation. Code and test pointers are collected
in the final section so the prose stays about the method.

## 1. The problem, and why one function serves three front ends

A researcher comparing treated and control groups across several covariates wants
a single number that says whether the groups are jointly balanced, and a p-value
for it that does not assume large samples. The natural statistic is a
Mahalanobis-type distance of the treated-group score vector from its permutation
mean,
$$
Q = (T - \mu)' \, M^{-1} \, (T - \mu),
\qquad T_c = \sum_i z_i \, \text{score}_{i,c},
$$
where $z$ is the 0/1 treatment vector, $T$ is the vector of treated-group score
sums across $r$ representations (covariates, basis functions, or test statistics),
$\mu = E[T]$ is the exact permutation mean, and $M$ is a fixed symmetric
positive-definite metric. The null is the within-stratum permutation of $z$ that
holds each stratum's treated count fixed.

Three different tests are the same computation with a different $M$:

- the `fastperm` omnibus combination uses $M = \Sigma$, a pooled covariance;
- the `RItools` `balanceTest` omnibus $d^2$ of Hansen and Bowers (2008) uses
  $M = V_d$, the permutation covariance of $T$, so $Q$ is asymptotically
  $\chi^2_{\,\mathrm{rank}(V_d)}$;
- the `RItools` covariate-covariance omnibus uses $M = \Sigma_x$, so $Q$ is
  asymptotically a weighted sum of chi-squares.

The point that makes this one function and not three: $M$ does not change across
the permutation. Both $V_d$ and $\Sigma_x$ are functions of the covariates and the
stratum sizes, not of $z$. So $M$ is a parameter of the computation, supplied by
the caller, and the metric is never hard-wired.

## 2. The rotation reduction: the metric is preprocessing

The first decision removes the metric from the hard part of the problem. For any
matrix $R$ with $R R' = M^{-1}$ (take $R = M^{-1/2}$ from the spectral
decomposition, dropping numerically-zero eigenvalues so a rank-deficient metric
projects onto its range),
$$
Q = (T-\mu)' M^{-1} (T-\mu) = (T-\mu)' R R' (T-\mu) = \lVert R'(T-\mu) \rVert^2 .
$$
So $Q$ is a plain sum of squares of the rotated, centred statistic
$e = R'(T-\mu)$. The metric enters only as the one-time rotation; everything
downstream works on $e$. This is exact, not an approximation: the test suite
checks it to machine precision against full enumeration of the permutation orbit.

In rotated coordinates write $e = T_y - \mu_y$ with $T_y = R'T$ the rotated
statistic and $\mu_y = R'\mu$ its mean. The eigenvalues $\lambda_1, \dots,
\lambda_r$ of the rotated permutation covariance $R' V_d R$ are the weights that
appear below.

## 3. The two engines

`fastperm_spa_quadratic(scores, treatment, strata, metric, method)` offers two
inversions of the same statistic.

- `method = "gaussian"` (M1) treats $e$ as Gaussian with the exact permutation
  covariance. Then $Q$ is a weighted sum of independent $\chi^2_1$ variables,
  $Q = \sum_k \lambda_k \chi^2_{1}$, whose cumulant generating function (CGF) is
  closed form and inverted by Lugannani-Rice.
- `method = "saddlepoint"` (M2) drops the Gaussian assumption and uses the
  *exact* permutation CGF of $Q$, so the true skewness and higher cumulants of the
  permutation law enter the tail.

M1 is fast and is the right tool when $e$ is close to Gaussian (large strata). M2
is the tool when $e$ is skewed and discrete -- few strata, small $n$, many
covariates -- which is exactly where balance testing lives and where the Gaussian
tail drifts.

## 4. M1: the Gaussian weighted-chi-square tail

### 4.1 The CGF and Lugannani-Rice

For $Q = \sum_k \lambda_k \chi^2_1$ with $\lambda_k > 0$ the CGF is
$$
K(\theta) = -\tfrac12 \sum_k \log(1 - 2\theta\lambda_k),
$$
analytic and steep on $\theta \in (-\infty,\, 1/(2\lambda_{\max}))$, with a pole
at $\theta = 1/(2\lambda_{\max})$. The Lugannani-Rice tail approximation is
$$
P(Q \ge q) \approx 1 - \Phi(\hat w) + \phi(\hat w)\left(\frac{1}{\hat u} - \frac{1}{\hat w}\right),
\quad
\hat w = \operatorname{sign}(\hat\theta)\sqrt{2(\hat\theta q - K(\hat\theta))},
\quad
\hat u = \hat\theta\sqrt{K''(\hat\theta)},
$$
where the saddlepoint $\hat\theta$ solves $K'(\hat\theta) = q$. This reproduces
`CompQuadForm::imhof` (the exact inversion of the same Gaussian-$d$ distribution)
to about 1% relative error in the tail -- the saddlepoint's own accuracy, not
slack -- and reduces to the $\chi^2_r$ tail when $M = V_d$ makes all
$\lambda_k = 1$.

### 4.2 The near-mean correction

The term $1/\hat u - 1/\hat w$ is a removable $0/0$ at $q = E[Q]$, where
$\hat w \to 0$ and $\hat u \to 0$. Worse, the difference loses precision to
cancellation ($\hat w \approx \hat u$) over a whole neighbourhood of the mean, not
just at the point. A naive guard that returns the normal tail there drops the
third-cumulant (skewness) term and, over the band where the raw formula is noisy,
makes the estimated tail non-monotone in $q$.

The fix replaces the term by its Taylor series in $\theta$ near the mean,
$$
\frac{1}{\hat u} - \frac{1}{\hat w} = c_0 + c_1 \theta + O(\theta^2),
\quad
c_0 = -\frac{K'''(0)}{6\,K''(0)^{3/2}},
\quad
c_1 = \frac{5\,K'''(0)^2}{24\,K''(0)^{5/2}} - \frac{K''''(0)}{8\,K''(0)^{3/2}},
$$
with the central cumulants $K''(0) = 2\sum\lambda_k^2$,
$K'''(0) = 8\sum\lambda_k^3$, $K''''(0) = 48\sum\lambda_k^4$. The leading term
gives the Daniels (1987) limit
$P(Q \ge E[Q]) = \tfrac12 + \phi(0)\,c_0
= \tfrac12 - K'''(0)/(6\sqrt{2\pi}\,K''(0)^{3/2})$,
which is below $1/2$ for a right-skewed $Q$. The series is used while
$|\hat w| < 0.1$, where the raw difference is unreliable; the direct formula runs
in the tails, where it is accurate. This region carries $p \approx 0.5$ and never
affects a rejection decision, but the function is now monotone and skewness-aware
there rather than flat.

## 5. M2: the exact permutation CGF for non-Gaussian $d$

### 5.1 The obstacle

The Lugannani-Rice machine needs only the CGF of the scalar $Q$,
$K_Q(\theta) = \log E[e^{\theta Q}]$. For Gaussian $e$ that CGF is closed form
(Section 4). For the true, non-Gaussian permutation law of $e$ there is generally
no closed form for $K_Q$, and supplying it is the entire problem. Every standard
quadratic-form tail method (Imhof, Davies, Farebrother, Liu-Tang-Zhang, and the
saddlepoint version Kuonen) reduces $Q$ to a weighted sum of chi-squares, which is
a Gaussian-only fact; none of them applies to a non-Gaussian, lattice $e$.

### 5.2 The Hubbard-Stratonovich identity

We do have the *joint* CGF of $e$, $K_e(s) = \log E[e^{s'e}]$, in closed form: it
is the elementary-symmetric permutation CGF evaluated on the scores projected onto
$s$. The Hubbard-Stratonovich identity converts the quadratic form into a Gaussian
average of that linear CGF. For $\theta > 0$,
$$
e^{\theta \lVert e \rVert^2}
= E_{W \sim N(0, I_r)}\!\left[\, e^{\sqrt{2\theta}\, W'e} \,\right],
$$
because $E_W[e^{a'W}] = e^{\lVert a\rVert^2/2}$ with $a = \sqrt{2\theta}\,e$. Taking
the expectation over $e$ inside (the dependence on $e$ is now linear) gives the
moment generating function of $Q$ as an exact $r$-dimensional Gaussian integral,
$$
M_Q(\theta) = E_{W \sim N(0, I_r)}\!\left[\, \exp\!\big(K_e(\sqrt{2\theta}\, W)\big) \,\right],
\qquad K_Q(\theta) = \log M_Q(\theta).
$$
Because $e$ has finite (permutation) support, $K_e$ grows at most linearly, the
$-\lVert W\rVert^2/2$ Gaussian weight dominates, and the integral converges for all
$\theta$. The bounded support also means $Q$ is bounded, so $K_Q$ is entire: it has
no pole, unlike the Gaussian weighted-chi-square CGF.

### 5.3 Why Laplace fails and quadrature succeeds

A natural first idea is to evaluate the Gaussian integral by Laplace's method
(steepest descent on $W$). It does not work, for an instructive reason. The CGF
$K_e$ is centred, so $\nabla K_e(0) = 0$ and $W = 0$ is always the critical point;
the leading-order Laplace expansion there returns exactly the Gaussian $K_Q$ and
discards all of the non-normality. The non-Gaussian content lives in the full
integral, not in its mode. (The same fact has a clean reading through
Kolassa-Robinson, below: a steepest-descent evaluation re-imposes a degenerate
surface approximation; doing the angular integral exactly is what keeps the
non-normality.)

The fix is to evaluate the Gaussian average by Gauss-Hermite quadrature -- the
standard rule for a Gaussian expectation -- rather than by Laplace. With a tensor
product of one-dimensional Gauss-Hermite nodes mapped to the $N(0, I_r)$ measure,
$$
K_Q(\theta) = \log \sum_k e^{\,\ell_k}\, \exp\!\big(K_e(\sqrt{2\theta}\, z_k)\big),
$$
with nodes $z_k$ and log-weights $\ell_k$. This recovers the exact $K_Q$: against
full enumeration of the orbit, the Gauss-Hermite value matches to about $10^{-15}$
for moderate $\theta$ and to about $10^{-6}$ in the deep tail.

### 5.4 Inversion, derivatives, and the higher-order option

With $K_Q$ in hand the inversion is the same Lugannani-Rice as M1. Because the
Gauss-Hermite $K_Q$ is accurate to near machine precision, its $\theta$-derivatives
$K_Q'$ and $K_Q''$ are taken by finite differences -- no analytic gradient or
Hessian of $K_e$ is needed. The saddlepoint $\hat\theta$ solves
$K_Q'(\hat\theta) = q$; since $Q$ is bounded there is no pole to bracket against,
and $K_Q'$ maps $\theta$ to the support range of $Q$.

Alongside the Lugannani-Rice tail, M2 returns the Barndorff-Nielsen $r^*$
refinement,
$$
P(Q \ge q) \approx 1 - \Phi\!\left(\hat w + \frac{1}{\hat w}\log\frac{\hat u}{\hat w}\right),
$$
which upgrades the relative error from $O(n^{-1})$ to $O(n^{-3/2})$ from the same
$\hat w, \hat u$. This is the one place a higher-order asymptotic enters the
pipeline; everything before it (the identity, the quadrature) is exact or
numerically exact.

### 5.5 Robustness and the dimension limit

M2 is built to return a finite, sensible number in every case. Below $E[Q]$, where
the upper-tail saddlepoint would need $\theta < 0$ (outside the
Hubbard-Stratonovich range), it returns the Gaussian-$d$ tail, which is accurate
there and not a rejection region. At the support edge or on a coarse lattice, where
the finite-difference second derivative goes unreliable, it again falls back to the
Gaussian tail with a warning rather than returning `NaN` or erroring.

The cost of the tensor Gauss-Hermite rule is $\text{nodes}^r$, so it is feasible
only for small $r$. At the default 32 nodes the function refuses $r \ge 4$ with an
error that points at the unbuilt sparse-grid / quasi-Monte-Carlo engine (M3), which
is the path to the many-covariate balance test rather than optional polish.

## 6. Where this sits in the literature

The nearest developed precedent is Kolassa and Robinson (2011), who give a
saddlepoint treatment of a multivariate permutation test. They do *not* approximate
the quadratic-form tail: they replace the quadratic form with a likelihood-ratio-
like statistic, because the direct large-deviation tail of a quadratic form in a
multivariate mean has relative error only $O(n^{-1/4})$ (Osipov, 1981), and they
sidestep that barrier by changing the statistic. Their own table shows the classical
Mahalanobis $D^2$ tail calibrated by chi-square is badly off in the tail, and they
state accurate tails for it are "not available." That classical $D^2$ is exactly
the statistic `RItools` and Hansen-Bowers report, and the cell M2 fills.

The $O(n^{-1/4})$ barrier is a property of the method, not of the statistic: it
comes from integrating a multivariate Edgeworth correction over a curved
(ellipsoidal) level set, where the surface contribution carries an extra factor of
$n^{1/2}$ (Robinson, Hoglund, Holst and Quine, 1990, Corollary 2.1). M2 never forms
that surface integral; the Hubbard-Stratonovich average runs over all directions of
$W$, which is the angular integration over the degenerate sphere of dominating
points, and Gauss-Hermite does it exactly. So the curvature is folded exactly into
the scalar $K_Q$ and the barrier does not arise.

Two honest caveats temper any rate claim, and we have now tested both. First,
escaping the curvature artifact is not the same as proving scalar $O(1/n)$: the
residual question is the discreteness of $Q$ at small stratum counts, not curvature.
Second, the honest competitor is Imhof, not Osipov's method; as the stratum count
grows the inner Laplace becomes exact and Imhof is itself $O(1/B)$, so M2's gain
over Imhof is a better *constant* at small $B$ -- carrying the true skewness and
kurtosis -- not a better rate.

The decisive check is an experiment: the relative error of M2 against the exact
permutation tail as the stratum count $B$ grows. We ran it (matched triples, skewed
$d$, metric $V_d$, full $3^B$ enumeration, tail depth $p = 0.05$, averaged over
random designs), and it confirms the second caveat against the first reading. M2's
relative error does *not* fall like $1/B$. It sits at 3 to 7 percent over $B = 8$ to
$12$ and, if anything, grows toward a floor, while the crude chi-square M1 keeps
falling toward its own $+0.5\%$ limit. So the simulation revises the optimistic
reading into the better-constant claim: M2 cuts the M1 p-value error 4- to 10-fold at
the stratum counts a balance test uses (escaping the useless Osipov-rate $D^2$ tail),
but it does not buy a better asymptotic rate.

The residual is a finite-discrete-representation effect, and a chain of controls
rules out the obvious corrections. It is not a pipeline artifact: the Gauss-Hermite
CGF reproduces the exact orbit CGF to $10^{-14}$ at the saddle, so the gap is genuine
Lugannani-Rice error inverting the exact CGF of a *finite* support. It is not the
continuous saddlepoint: the analytic chi-square LR error is $+0.5\%$ for $r = 1, 2, 3$
alike. It is not curvature: the bias *falls* with dimension -- $+15\%$ at $r = 1$,
$+7\%$ at $r = 2$, $+2\%$ at $r = 3$ -- where a count of discrete points against the
curved level set would grow with $r$. And it is not generic permutation discreteness:
the scalar sum saddlepoint on the same orbit is accurate to under one percent. What
remains is the gap between continuous Lugannani-Rice and the exact tail of a finite,
discrete representation of $Q$; a uniform law on $n$ i.i.d. exact chi-square draws
shows the same bias against an analytic LR error of $+0.5\%$. The practical reading is
favorable: the Mahalanobis omnibus is used at $r \ge 2$ (a single covariate is a
scalar test), the bias there is small and *falls* as covariates are added, and it is
conservative -- M2 reports p-values slightly too large. The full chain of
experiments is in `dev/route-b-rate-experiment-findings.md`; a fuller literature
discussion is in `dev/route-b-literature.md`.

On higher-order asymptotics specifically: the scalar $r^*$ of Section 5.4 is the
tail refinement of a statistic with a known CGF (Barndorff-Nielsen 1986;
Brazzale, Davison and Reid 2007 is the practical reference). It is distinct from
Brazzale's core program, the $r^*$ for a parameter of interest in a parametric
likelihood with nuisance parameters, which our randomization null -- no likelihood,
no nuisance parameter -- does not invoke.

## 7. Known limitation: coarse lattices and the finite-discrete residual

M2 is much more accurate than M1 on the dense, multi-stratum support balance
testing has -- 4 to 10 times better in the rate experiment of Section 6 -- but it
carries the finite-discrete residual diagnosed there, and a separate coarse-lattice
edge case. On a single or double stratum $V_d$ is proportional to the sample
covariance of the scores (so the metric choices coincide), and $Q$'s support is a
coarse, atom-heavy lattice; the continuous saddlepoint then under- or overshoots the
discrete tail, so M2 falls back to the Gaussian tail there. That fallback ships.

The dense-support residual (Section 6) is not addressed by a continuity correction,
and we are explicit that we do not implement one: the controls show it is not a
regular lattice (no fixed span; the local span shrinks like $1/N$, which would
predict a vanishing bias the data contradict), not the empirical CGF's tail-max, not
boundary curvature, and not the continuous chi-square saddlepoint. It is the gap
between continuous Lugannani-Rice and the exact tail of a finite, discrete $Q$, and
it shrinks with dimension rather than growing. Because that residual is small and
falling at the $r \ge 2$ where the omnibus is used, conservative in sign, and already
far below M1, a finite-discrete correction (an Euler-Maclaurin-type boundary term) is
research-level work whose payoff we judge not worth its complexity for this
application. The path forward is to wire M2 into the balance-test front ends, not to
chase the last few percent at low $r$.

## 8. Validation

Every claim above is checked against the exact permutation distribution, obtained
by walking the entire within-stratum orbit (no Monte Carlo error) for the small
designs the tests use. The oracle is `enumerate_Q` in the test helper. On a dense
ten-pair orbit with skewed scores and $M = \Sigma_x$, M2 tracks the exact tail to
within 3 to 5 percent where the Gaussian M1 is off by 15 to 340 percent -- the gap
widens with the skewness of $e$, as the theory predicts.

```{r example, eval = TRUE}
library(fastperm)
# eight matched pairs; the second covariate is skewed (fibonacci), so the
# permutation law of Q is non-normal and the Gaussian tail drifts
scores <- cbind(rep(c(1, 2), 8) + rep(0:7, each = 2) * 0.4,
                c(rbind(rep(1, 8), c(2, 3, 5, 8, 13, 21, 34, 55))))
strata <- rep(1:8, each = 2)
# treat the high-score unit in five of the eight pairs (a moderate imbalance)
z <- rep(0, 16)
for (b in 1:8) z[if (b <= 5) 2 * b else 2 * b - 1] <- 1
# Gaussian (M1) versus the non-normal permutation saddlepoint (M2)
fastperm_spa_quadratic(scores, z, strata, method = "gaussian")$p.value
res <- fastperm_spa_quadratic(scores, z, strata, method = "saddlepoint")
c(lugannani_rice = res$p.value, r_star = res$p.value.rstar)
```

## 9. Decisions, with code and test locations

Each row is a decision, the function that implements it (in
`R/saddlepoint-quadratic.R` unless noted), and the test that guards it (in
`tests/testthat/test-saddlepoint-quadratic.R` unless noted).

- **Rotation reduction $Q = \lVert R'(T-\mu)\rVert^2$, metric as preprocessing.**
  `.quad_inv_sqrt` (the $M^{-1/2}$ rotation). Test: "the rotation reduction
  $Q = \lVert d_y\rVert^2$ is an exact identity".
- **Metric as a caller-supplied parameter ("cov" or a matrix).**
  `fastperm_spa_quadratic` argument `metric`. Tests: "M1 with metric = 'cov'
  reduces to the chi-square tail"; "M1 reproduces imhof for a non-identity metric".
- **Exact permutation mean and covariance (Strasser-Weber).**
  `fastperm_linear_cgf_mv` in `R/cgf-stratified-mv.R`. Test: "the enabler's
  permutation covariance matches the orbit covariance".
- **M1 Lugannani-Rice on the weighted-chi-square CGF.** `.quad_lr_upper`. Tests:
  "M1 with metric = 'cov' reduces to the chi-square tail"; "M1 reproduces imhof for
  a non-identity metric".
- **M1 near-mean Daniels limit and Taylor series (the cancellation fix).**
  `.quad_lr_upper` (coefficients `c0`, `c1`). Tests: ".quad_lr_upper returns the
  Daniels skewness limit at the mean, not 1/2"; ".quad_lr_upper is continuous and
  monotone across the mean".
- **M2 Hubbard-Stratonovich CGF by Gauss-Hermite quadrature.** `.make_KQ_quad`,
  `.tensor_normal`, `.gh_nodes`. The Laplace dead-end and the quadrature fix are
  recorded in `dev/route-b-quadratic-inversion.md`.
- **Vectorised multi-node elementary-symmetric evaluation (the speedup).**
  `.log_esym_cols` (a column-vectorised `.log_esym` from `R/cgf-stratified.R`).
- **M2 finite-difference $\theta$-derivatives (justified by machine-precision GH).**
  `.quad_spa_upper`.
- **M2 Lugannani-Rice tail plus Barndorff-Nielsen $r^*$.** `.quad_spa_upper`
  (returned as `p.value` and `p.value.rstar`). Test: "M2 (saddlepoint) returns a
  finite tail and a Barndorff-Nielsen r*".
- **M2 Gaussian fallback below $E[Q]$, at the support edge, and on coarse
  lattices.** `.quad_spa_upper`. Test: "M2 is never worse than M1 against the exact
  tail at the observed Q".
- **M2 accuracy and improvement over M1 on the skewed dense case.**
  Test: "M2 matches the exact permutation tail and beats M1 where d is skewed".
- **Tensor-grid dimension guard pointing at M3.** `fastperm_spa_quadratic`
  (the `nodes^r > 250000` check). Test: "M2 errors for rank too large for the
  tensor grid, pointing past M2".
- **Exact-enumeration oracle (ground truth for all of the above).**
  `enumerate_Q`, `enum_quadratic_tail` in
  `tests/testthat/helper-quadratic-enumerate.R`.

## References

- Barndorff-Nielsen, O. E. (1986). Inference on full or partial parameters based
  on the standardized signed log likelihood ratio. *Biometrika* 73, 307-322.
- Brazzale, A. R., Davison, A. C. and Reid, N. (2007). *Applied Asymptotics: Case
  Studies in Small-Sample Statistics*. Cambridge University Press.
- Daniels, H. E. (1987). Tail probability approximations. *International
  Statistical Review* 55, 37-48.
- Hansen, B. B. and Bowers, J. (2008). Covariate balance in simple, stratified and
  clustered comparative studies. *Statistical Science* 23, 219-236.
- Kolassa, J. and Robinson, J. (2011). Saddlepoint approximations for likelihood
  ratio like statistics with applications to permutation tests. *Annals of
  Statistics* 39, 3357-3368.
- Kuonen, D. (1999). Saddlepoint approximations for distributions of quadratic
  forms in normal variables. *Biometrika* 86, 929-935.
- Lugannani, R. and Rice, S. (1980). Saddle point approximation for the
  distribution of the sum of independent random variables. *Advances in Applied
  Probability* 12, 475-490.
- Osipov, L. V. (1981). On large deviations for sums of random vectors in
  $R^k$. *Journal of Multivariate Analysis* 11, 115-126.
- Robinson, J., Hoglund, T., Holst, L. and Quine, M. P. (1990). On approximating
  probabilities for small and large deviations in $R^d$. *Annals of Probability*
  18, 727-753.
